Auto-tuned Jacobi-like Multi-grid Smoother\\ for Fast Pressure Projection

### 2021-10-01 tex/article.tex

Pressure projection is a bottleneck in high-speed unsteady incompressible flow solvers. The pressure projection step typically makes up 25-50\% of the solution cost even in serial simulations, and the proportional cost grows in massively parallel simulations as the elliptic pressure Poisson equation requires communication throughout the computational domain instead of being confined to a local characteristic. Methods such as artificial-compressibility, smooth-particle-hydrodynamics, and particle-in-cell attempt to side-step this computational cost by modelling or otherwise under-resolving the pressure evolution compared to the fluid's momentum. However, these approaches lead to significant errors in pressure forces, thereby making them unsuitable for many applications or requiring explicit pressure corrections.

@@

However, the pressure projection in these all of these studies have not been augmented, leaving their acceleration as a nearly untouched topic. Some recent work has focused on the pressure projection step explicitly. Chaotic projection methods for massively parallel systems; not data driven. Multi-grid data-driven decomposition; high errors and unclear speed-up. This work links develops a high-speed parameterized smoother which outperforms classic smoothing methods on eleven 2D and 3D benchmarks. Moreover, as the approach in linear in the pressure and scale-invariant, the parameters can be tuned on a coarse simulation and maintain their accelerated performance on the fine simulation.

@@

The discrete Poisson equation at the heart of the projection step is defined as
\begin{equation}\label{eq:axb}
    A x = b
\end{equation}
where $x$ is the desired pressure field vector, $b$ is the source term (proportional to the divergence of the velocity field to be projected), and $A$ is the Poisson matrix. For conservative Poisson equations, $A$ is symmetric, has zero-sum rows and columns, is negative semi-definite, and has a single zero eigenvalue. The matrix is also extremely sparse. Using a second-order scheme on structured grids in $M$ dimensions, $A$ has only $M$ non-zero sub-diagonals. While the solution vector size $N$ may easily be larger than $10^6$, the matrix-vector multiplication $Ax$ only has computational cost $O(NM)=O(N)$ instead of $O(N^2)$.

Iterative methods solve equation \ref{eq:axb} by updating an approximate solution $x^k$ to a new solution $x^{k+1}$ with reduced error. As the problem is linear, the equation for the update $\epsilon$ is simply
\begin{equation}
    A \epsilon^k = r^k \equiv b - Ax^k
\end{equation}
where $r$ is the residual. In practise, only an approximate solution for $\epsilon$ is obtained, after which the solution and residual are incremented
\begin{equation}
    x^{k+1} = x^k+\epsilon^k, \quad r^{k+1} = r^k-A\epsilon^k.
\end{equation}
This process is iterated until the residual is reduced sufficiently, the required tolerance level being highly dependant on the application.

@@

The linearity of the system is a critical property which the data-driven method must maintain, but this linearity is only with respect to the update and residual fields. The coefficients of the ``MG network'' can all potentially be optimized for a specific sub-class of problems, and they can be made nonlinear functions of $A$. However, neither the restriction or prolongation operations are functions of $A$, and a workable method of including the dependence is unclear.\footnote{For example, the residual restriction could use $a_{ij}$ to define a non-uniform weight function $w_{ij} = r_i/r_j$, but since the matrix is symmetric the same weight should apply to $r_j/r_i$, resulting in a uniform pooling.} Similarly, a simple Jacobi preconditioner
\begin{equation}
    \epsilon = D^{-1}r
\end{equation}
where $D$ is the diagonal of $A$, is as fast as possible and is sufficient for the V-cycle to distribute the residual throughout the domain. Therefore, for the remainder of the paper we will focus on the smoother as the best candidate for acceleration, using a simple Jacobi preconditioner and uniform pooling/distribution for restriction/prolongation.

@@

There are few constraints on $\tilde A^{-1}$: it must have units of $A^{-1}$, and symmetry of $A$ implies the approximate inverse should also be symmetric. For simplicity, the diagonal and off-diagonal coefficients of $\tilde A^{-1}$ are all constructed independently
\begin{equation}
    \tilde a^{-1}_{ii} = \frac{f_D(a_{ii}/s\,|\theta_D)}{a_{ii}} , \quad
    \tilde a^{-1}_{ij} = \frac{f_L(a_{ij}/s\,|\theta_L)}{a_{ii}+a_{jj}}
\end{equation}
where function inputs are scaled by the maximum off-diagonal $s=\max(A-D)$ and the outputs are scaled by the diagonal elements. Note a Jacobi smoother would use $f_D=1$ and $f_L=0$.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{2D-box}
        \label{fig:2D-box}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{2D-$\mu$}
        \label{fig:2D-mu}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{2D-sphere}
        \label{fig:2D-sphere}
    \end{subfigure}
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/crossloss.png}
        \caption{$\log_{10}$ residual reduction}
        \label{fig:cross plot}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/crosscount.png}
        \caption{Solver time}
        \label{fig:synthetic time}
    \end{subfigure}
    \caption{(a-c) Sample solutions from the data-sets of three of the six synthetic cases. (d) Residual reduction over a single Multi-grid V-cycle on the `test' cases after tuning using the `train' case. Training case `union' refers to the smoother trained on all of the synthetic cases. (b) Time to reduce pressure residual by $10^{-3}$ for classical and parameterized smoothers on each synthetic case. Time is relative to the time of a single V-cycle using the Jacobi smoother.}
    \label{fig:synthetic cases}
\end{figure}

@@

A set of six synthetic projection cases were constructed to establish the characteristics of the parameterized smoother. The three 2D cases are show in Figure~\ref{fig:2D-box}-\ref{fig:2D-sphere}, and each has a matching 3D case. The `box' cases have a simple constant gradient solution throughout the domain, with the direction and magnitude randomized. The dipole `$\mu$' cases have a localized dipole within a quiescent flow with randomized placement, direction and magnitude. The `sphere' cases have a moving sphere immersed within a quiescent flow with randomized radius, location, and velocity direction. Neumann boundary conditions are applied on all boundaries, including the immersed sphere, through adjustment of the $A$ matrix coefficients which are otherwise uniform. All cases are initialized with $x^0=0$, and all cases feature initially localized residuals; i.e. $r^0\ne0$ only in a fraction of the cells, but $x$ must be updated throughout the domain, starkly illustrating the elliptic nature of the projection step.

Figure~\ref{fig:cross plot} shows the `test' performance of the parameterized smoother over a single V-cycle on 100 randomized examples after optimization using 100 examples of the `test' case. Each example uses a grid size of n=32 point in each direction ($N=32^2$ in 2D and $N=32^3$ in 3D) as testing with other resolutions found essentially no change in the performance. As expected, the performance is best when testing and training on the same case, with the residual reduced by $10^{-1.18}$ on the 2D-sphere case down to $10^{-1.75}$ on the 2D-box case. The performance for the simplest `box' and 2D-$\mu$ cases generalizes poorly, with some cases diverging instead converging, but the optimized 2D-sphere smoother generalizes essentially as well as the `union' smoother trained on all of the data. 
%The performance of the `union' smoother is only 1-4\% lower than the smoother trained on the test case for the dipole and sphere cases.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{circle flow}
        \label{fig:circle}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{Taylor-Green Vortex}
        \label{fig:TGV}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{flapping wing}
        \label{fig:donut}
    \end{subfigure}
    \begin{minipage}{0.45\textwidth}
        \begin{subfigure}{\textwidth}
        \centering
        \caption{torus flow}
        \label{fig:donut}
        \end{subfigure}
        \begin{subfigure}{\textwidth}
        \centering
        \caption{swimming shark}
        \label{fig:donut}
        \end{subfigure}
    \end{minipage}
    \hfill
    \begin{subfigure}[b]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/compareloss.png}
        \caption{$\log_{10}$ residual reduction}
        \label{fig:synthetic time}
    \end{subfigure}
    \caption{(a-e) Snapshots from the five flow simulation cases. The Taylor-Green Vortex and torus flow are 3D cases, and the flapping wing and swimming shark have dynamic geometries. (f) Residual reduction over a single Multi-grid V-cycle for classical and parameterized smoothers on each case. The `transfer' smoother has been trained on the union of the synthetic data sets from Figure~\ref{fig:synthetic cases}.}
    \label{fig:simulation cases}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/scaleloss.png}
        \caption{$\log_{10}$ residual reduction}
        \label{fig:scaled loss}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/crosscount.png}
        \caption{Solver time}
        \label{fig:simulation time}
    \end{subfigure}
        \caption{(a) Residual reduction over a single Multi-grid V-cycle on the full-resolution case. The `transfer' smoother has been trained on the union of the synthetic data sets from Figure~\ref{fig:synthetic cases}. The $\frac 18, \frac 14, \frac 12$ smoothers have been trained on simulations with the indicated reduced resolution \textit{in each spacial and temporal dimension}. (b) Time to reduce pressure residual by $10^{-3}$ for classical and parameterized smoothers on each simulation case. Time is relative to the time of a single V-cycle using the Jacobi smoother.}
        \label{fig:tuned simulation}
\end{figure}

>>>>

Pressure projection is a bottleneck in high-speed unsteady incompressible flow solvers. Constructing approximate solutions the discrete pressure Poisson system typically makes up 25-50\% of the solution cost even in serial simulations, and the proportional cost grows in massively parallel simulations as the elliptic equation requires communication throughout the computational domain instead of being confined to a local characteristic. Methods such as artificial-compressibility, smooth-particle-hydrodynamics, and particle-in-cell attempt to side-step this computational cost by modelling or otherwise under-resolving the pressure evolution compared to the fluid's momentum. However, these approaches lead to significant errors in pressure forces, thereby making them unsuitable for many applications or requiring explicit pressure corrections.

@@

However, none of these studies have applied data-driven techniques to accelerate the critical projection step itself. Some recent work has focused on the pressure projection step explicitly. Chaotic projection methods for massively parallel systems; not data driven. Multi-grid data-driven decomposition; high errors and unclear speed-up. This work links develops a high-speed parameterized smoother which outperforms classic smoothing methods on eleven 2D and 3D benchmarks. Moreover, as the approach in linear in the pressure and scale-invariant, the parameters can be tuned on a coarse simulation and maintain their accelerated performance on the fine simulation.

@@

The discrete Poisson equation in the projection step is defined as
\begin{equation}\label{eq:axb}
    A x = b
\end{equation}
where $x$ is the desired pressure field vector, $b$ is the source term (proportional to the divergence of the velocity field to be projected), and $A$ is the Poisson matrix. For conservative Poisson equations, $A$ is symmetric, has zero-sum rows and columns, is negative semi-definite, and has a single zero eigenvalue. The matrix is also extremely sparse. Using a second-order scheme on structured grids in $M$ dimensions, $A$ has only $M$ non-zero sub-diagonals. While the solution vector size $N$ may easily be larger than $10^6$, the matrix-vector multiplication $Ax$ only has computational cost $O(NM)=O(N)$ instead of $O(N^2)$.

Iterative methods solve equation \ref{eq:axb} by updating an approximate solution $x^k$ to a new solution $x^{k+1}$ with reduced error. As the problem is linear, the equation for the update $\epsilon$ is simply
\begin{equation}\label{eq:aer} 
    A \epsilon^k = r^k \equiv b - Ax^k
\end{equation}
where $r$ is the residual. In practise, only an approximate solution for $\epsilon$ is obtained, after which the solution and residual are incremented
\begin{equation}
    x^{k+1} = x^k+\epsilon^k, \quad r^{k+1} = r^k-A\epsilon^k.
\end{equation}
This process is iterated until the residual is reduced sufficiently, the required tolerance level being highly dependant on the application.

@@

The linearity of equation \ref{eq:aer} is a critical property which the data-driven method must maintain in order to be useful in practise, but this linearity is only with respect to the update and residual fields. The operators of the ``MG network'' can all potentially be optimized for a specific sub-class of problems. In particular, those operators can be made into parameterized nonlinear functions of $A$, embedding information about the discrete problem and boundary conditions, without violating linearity.

Which MG operation is the best candidate for such an approach? The prolongation operator (restriction simply being its transpose) is not inherently a function of $A$, and while related super-resolution operators have seen improvements in the literature, these are always nonlinear functions of the field. The precondition operator is an option, but a simple Jacobi preconditioner
\begin{equation}
    \epsilon = D^{-1}r
\end{equation}
where $D$ is the diagonal of $A$, is as fast as possible and is sufficient for the V-cycle to distribute the residual throughout the domain. Therefore, for the remainder of the paper we will focus on the smoother as the best candidate for acceleration, using a simple Jacobi preconditioner and uniform pooling/distribution for restriction/prolongation.

@@

There are few constraints on $\tilde A^{-1}$: it must have units of $A^{-1}$, and symmetry of $A$ implies the approximate inverse should also be symmetric. For simplicity, the diagonal and off-diagonal coefficients of $\tilde A^{-1}$ are constructed independently 
\begin{equation}
    \tilde a^{-1}_{ii} = \frac{f_d(a_{ii}/s\,|\theta_d)}{a_{ii}} , \quad
    \tilde a^{-1}_{ij} = \frac{f_o(a_{ij}/s\,|\theta_o)}{a_{ii}+a_{jj}}
\end{equation}
where function inputs are scaled by the maximum off-diagonal $s=\max(A-D)$ and the outputs are scaled by the diagonal elements. Note a Jacobi smoother would use a diagonal function $f_d=1$ and off-diagonal $f_o=0$.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{2D-box}
        \label{fig:2D-box}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{2D-$\mu$}
        \label{fig:2D-mu}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{2D-sphere}
        \label{fig:2D-sphere}
    \end{subfigure}
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/crossloss.png}
        \caption{single-cycle residual reduction}
        \label{fig:cross plot}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/synthetic_timing.png}
        \caption{solver time}
        \label{fig:synthetic time}
    \end{subfigure}
    \caption{(a-c) Sample solutions from the data-sets of three of the six synthetic cases. (d) Residual reduction over a single Multi-grid V-cycle on the `test' cases after tuning using the `train' case. Training case `union' refers to the smoother trained on all of the synthetic cases. (b) Time to reduce pressure residual by $10^{-3}$ for classical and parameterized smoothers on each synthetic case. Time is relative to the time of a single V-cycle using the Jacobi smoother.}
    \label{fig:synthetic cases}
\end{figure}

@@

A set of six synthetic projection cases were constructed to establish the characteristics of the parameterized smoother. The three 2D cases are show in Figure~\ref{fig:2D-box}-\ref{fig:2D-sphere}, and each has a matching 3D case. The `static' cases have a simple hydro-static gradient solution throughout the domain, with the direction and magnitude randomized. The `dipole' cases have a localized dipole within a quiescent flow with randomized placement, direction and magnitude. The `sphere' cases have a moving sphere immersed within a quiescent flow with randomized radius, location, and velocity direction. Neumann boundary conditions are applied on all boundaries, including the immersed sphere, through adjustment of the $A$ matrix coefficients which are otherwise uniform. All cases are initialized with $x^0=0$, and all cases feature initially localized residuals; i.e. $r^0\ne0$ only in a fraction of the cells, but $x$ must be updated throughout the domain, starkly illustrating the elliptic nature of the projection step.

Figure~\ref{fig:cross plot} shows the `test' performance of the parameterized smoother over a single V-cycle on 100 randomized examples after optimization using 100 examples of the `test' case. Each example uses a grid size of n=32 point in each direction ($N=32^2$ in 2D and $N=32^3$ in 3D) as testing with other resolutions found essentially no change in the performance. As expected, the performance is best when testing and training on the same case, with the residual reduced by $10^{-1.18}$ on the 2D-sphere case down to $10^{-1.75}$ on the 2D-box case. While the performance for the simplest cases generalizes poorly, the 2D-sphere smoother generalizes essentially as well as the `union' smoother trained on all of the data. 
%The performance of the `union' smoother is only 1-4\% lower than the smoother trained on the test case for the dipole and sphere cases.

Figure~\ref{fig:synthetic time} characterizes the acceleration achieved by the data-driven smoother using the time taken to reduce the residual of each case by $10^{-3}$, scaled by the time to run a single V-cycle using a Jacobi smoother. The results for the Jacobi smoother are not given, since thousands of iterations are required. Remarkably, the Jacobi-like parameterized smoother is only 36\% slower per iteration, and yet converges in only a few V-cycles. This results in a 80-175\% speed up relative to optimized serial Gauss-Sidel and SOR smoothers, and the speed-up will be even more favorable in parallel computation.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{circle flow}
        \label{fig:circle}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{Taylor-Green Vortex}
        \label{fig:TGV}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \centering
        \caption{flapping wing}
        \label{fig:donut}
    \end{subfigure}
    \begin{minipage}{0.47\textwidth}
        \begin{subfigure}{\textwidth}
        \centering
        \caption{torus flow}
        \label{fig:donut}
        \end{subfigure}
        \begin{subfigure}{\textwidth}
        \centering
        \caption{swimming shark}
        \label{fig:donut}
        \end{subfigure}
    \end{minipage}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/compareloss.png}
        \caption{single-cycle residual reduction}
        \label{fig:simulation residual}
    \end{subfigure}
    \caption{(a-e) Snapshots from the five flow simulation cases. The Taylor-Green Vortex and torus flow are 3D cases, and the flapping wing and swimming shark have dynamic geometries. (f) Residual reduction over a single Multi-grid V-cycle for classical and parameterized smoothers on each case. The `transfer' smoother has been trained on the union of the synthetic data sets from Figure~\ref{fig:synthetic cases}.}
    \label{fig:simulation cases}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/diag_fun.png}
        \caption{diagonal functions}
        \label{fig:scaled loss}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/lower_fun.png}
        \caption{off-diagonal functions}
        \label{fig:simulation time}
    \end{subfigure}
        \caption{(a) Diagonal and (b) off-diagonal parameterized functions after optimization on each simulation case. The `transfer' smoother was tuned on the synthetic cases and `Jacobi' is shown for comparison.}
        \label{fig:tuned simulation}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/scaleloss.png}
        \caption{single-cycle residual reduction}
        \label{fig:scaled loss}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/crosscount.png}
        \caption{solver time}
        \label{fig:simulation time}
    \end{subfigure}
        \caption{(a) Residual reduction over a single Multi-grid V-cycle on the full-resolution case. The `transfer' smoother has been trained on the union of the synthetic data sets from Figure~\ref{fig:synthetic cases}. The $\frac 18, \frac 14, \frac 12$ smoothers have been trained on simulations with the indicated reduced resolution \textit{in each spacial and temporal dimension}. (b) Time to reduce pressure residual by $10^{-3}$ for classical and parameterized smoothers on each simulation case. Time is relative to the time of a single V-cycle using the Jacobi smoother.}
        \label{fig:tuned simulation}
\end{figure}
